Variance reduction for MCMC methods via martingale representations

### 2020-03-19 aap/Revision/vr-mcmc-revision-new.tex
update
lower bound

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N},
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}). In many situations the coefficients $\bar{a}_{r,k}$ also decay exponentially fast as $r$ increases. The following result can be easily established.
\begin{lem}
\label{lem:coef_decrease}
Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
\begin{equation}
|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
\end{equation}
where $\rho\in (0,1)$ and $C$ is an absolute constant.
\end{lem}

\section{Estimation of coefficients}
The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
(the so-called ``training paths''):
\[
\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
\]
with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n_{tr}-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{equation}
for \(r=1,\ldots, n_{tr}-1,\)
where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $n_tr$ is a fixed natural number. 
Next  we estimate the coefficients \( \bar a_{l,k}\) using
the formula
\begin{equation}
\label{eq:a_est_int}
\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
\end{equation}
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories. In many cases integration can be done in closed analytical form. For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\begin{eqnarray*}
\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\end{eqnarray*}
Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
\\
\leq \frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}
Yet it is impractical to use estimator of the form $\widehat M_{K,n}^N(f)$, since it may be too expensive to compute for $n$ large enough. Instead we introduce the truncated version of this control variate
\begin{eqnarray*}
\tilde M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\mathbbm{1}\{ |p-k| < n_{\mathrm{trunc}} \}\right]
\end{eqnarray*}
where $n_{\mathrm{trunc}} \in \nset$ is a hyperparameter. Validity of this approach follows from the following lemma

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{proof} Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Thus we represent the squared bias as 
\begin{eqnarray}
\label{eq:bias_1}
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=&\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
=
\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \leq \\
&\leq& 
2|\pi_{\gamma}(f) - \pi(f)|^2 + \frac{2}{n}\sum\limits_{k=N+1}^{N+n}\left[\PE f(X_k^x) - \pi_{\gamma}(f)\right]^2
\end{eqnarray}
 From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad |\PE f(X_k^x) - \pi_{\gamma}(f)| \leq C_2 \rho^n\|f\|_{V}V(x)
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{13} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$. Hence, combining \eqref{eq:bias_1} and \eqref{eq:bias_2}, we obtain \eqref{eq:sq_bias}.
\end{proof}

@@

@@

\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} \eqsp,

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

>>>>

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N},
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}). In many situations the coefficients $\bar{a}_{r,k}$ also decay exponentially fast as $r$ increases. The following result can be easily established.
\begin{lem}
\label{lem:coef_decrease}
Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
\begin{equation}
|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
\end{equation}
where $\rho\in (0,1)$ and $C$ is an absolute constant. In particular, if $f$ is uniformly bounded we can take $V(x)\equiv 1$ to get
\begin{eqnarray*}
|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{\infty}.
\end{eqnarray*}

\end{lem}
This lemma shows that in fact  only a logarithmic number (in $n$) of coefficients need to be computed.

\section{Estimation of coefficients and variance reduction}
The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions 
\[
S_{q}(x)=\mathsf{E}\left[\sum_{r=0}^q f(X^x_{r})\right]
\]
for $q=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
(the so-called ``training paths''):
\[
\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
\]
with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
\widehat{S}_q = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l=N+1}^{L}\left| \sum_{r = 0}^{q}  f(X_{r+l}^{(s)}) - \psi(X^{(s)}_l)\right|^2
\end{equation}
for some $L$ with $L+q\leq N+n,$
where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and set  $\widehat{S}_{R,q}=\max\{\widehat{S}_q,R\}$ for some $R>0.$
Next  we estimate the coefficients \( A_{q,k}\) using
the formula
\begin{equation}
\label{eq:a_est_int}
\widehat A_{q,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat S_{R,q}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
\end{equation}
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat S_{R,q}$ derived from training trajectories. In many cases integration can be done in closed analytical form. For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((S_{q}),\) then \((\widehat A_{q,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
Upon estimating  the functions \((\widehat A_{q,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\begin{eqnarray*}
\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\end{eqnarray*}
Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
\\
\leq \frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}
Combining this with \eqref{eq:expression-variance} results in 
\begin{multline*}
\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
\\
+\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right]
\end{multline*}

Suppose that $\|f\|_\infty\leq F$ and $\|A_{l,k}\|_{\infty}\leq R$ for all $k=1,\ldots,K,$ $l=1,2,\ldots $ and some constants $F,K>0.$ The standard regression estimates imply that 
\begin{eqnarray*}
\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right]&\lesssim &  \frac{(R^2 \vee F^2)\,\mathrm{Dim}(\Psi)}{T}
\\
&&+\sup_{\psi\in\Psi}\|A_{n-l,k}-\psi\|_{L^2(P^x_{N+l-1})}
\end{eqnarray*}
with high probability.
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
\begin{eqnarray*}
A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
\end{eqnarray*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{proof} Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Thus we represent the squared bias as 
\begin{eqnarray}
\label{eq:bias_1}
\nonumber
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=&\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
=
\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \\
&\leq& 
2|\pi_{\gamma}(f) - \pi(f)|^2 + \frac{2}{n}\sum\limits_{k=N+1}^{N+n}\left[\PE f(X_k^x) - \pi_{\gamma}(f)\right]^2
\end{eqnarray}
 From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad |\PE f(X_k^x) - \pi_{\gamma}(f)| \leq C_2 \rho^n\|f\|_{V}V(x)
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{lem:v_ergodicity} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$. Hence, combining \eqref{eq:bias_1} and \eqref{eq:bias_2}, we obtain \eqref{eq:sq_bias}.
\end{proof}

@@

Note that the bound of lemma~\ref{lem:variance} is sharp, namely, the following statement holds:
\begin{lem} 
\label{lem:ula_var_lower_bound}
Let $f(x) = x$ and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, with $\pi(x) = \frac{1}{\sqrt{2\pi}}e^{-\frac{x^2}{2}}$ and assume also $n\gamma > C_0$ for some fixed $C_0$. Then it holds
\begin{equation}
\label{eq:lower_bound}
\PVar[\pi_n^0(f)] \geq \frac{2}{n\gamma} - \mathcal{O}\left(\frac{1}{n^2\gamma^2}\right) \eqsp,
\end{equation}
\end{lem}

@@

\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} 
\eqsp,

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}

@@

\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K}A^2_{s,\mathbf{k}}(x)
